---
title: "Chapter 7"
format: html
---

```{r}
turtles = read.table("../data/PaintedTurtles.txt", header = TRUE)
turtles[1:4, ]
```


```{r}
load("../data/athletes.RData")
athletes[1:3, ]
```

```{r}
stopifnot(nrow(athletes) == 33, ncol(athletes) == 10)
athletesSummary = lapply(athletes, function(x) c(min = min(x), max = max(x), mean = mean(x), sd = sd(x)))
```

```{r}
load("../data/Msig3transp.RData")
round(Msig3transp,2)[1:5, 1:6]
```

```{r}
metab = t(as.matrix(read.csv("../data/metabolites.csv", row.names = 1)))
metab[1:4, 1:4]
```

```{r}
cor(turtles[, -1])
```




```{r}
library(ggplot2)
apply(turtles[,-1], 2, sd)
apply(turtles[,-1], 2, mean)
scaledTurtles = scale(turtles[, -1])
apply(scaledTurtles, 2, mean)
apply(scaledTurtles, 2, sd)
data.frame(scaledTurtles, sex = turtles[, 1]) |>
  ggplot(aes(x = width, y = height, group = sex)) +
    geom_point(aes(color = sex)) + coord_fixed()
```


```{r}
x1=1;x2=3;y1=1;y2=2;ax=2.5;ay=3;
df=data.frame(x=c(x1,x2,ax),y=c(y1,y2,ay))
ggplot(df, aes(x=x,y=y)) + 
       geom_point(size=2) +
       geom_abline(intercept=0.5,slope=0.5, color="red", linewidth=1.3) + 
       xlim(c(0,4)) + 
       ylim(c(0,4)) +
       geom_segment(x=x1,y=y1,xend=x2-0.5,yend=y2-0.25,arrow=arrow(length = unit(0.3,"cm")),color="blue") +
       geom_segment(x=ax,y=ay,xend=x2,yend=y2,arrow=arrow(length = unit(0.3,"cm")),color="orange",
                    linetype = 5, linewidth = 1.2, alpha = 0.5) + 
       annotate("text", x = ax+0.2, y = ay+0.15, label = "A", size=6) +
       annotate("text", x = x2, y = y2-0.5, label = "proj_v(A)", size=6) +
       annotate("text", x = x1+0.75, y = y1+0.24, label = "v", size=6, color="blue") +
       annotate("text", x = x1-0.2, y = y1+ 0.2, label = "O", size=6) +
       coord_fixed() + 
       theme_void() +
       geom_point(size=2)
```

```{r}
athletes = data.frame(scale(athletes))
ath_gg = ggplot(athletes, aes(x = weight, y = disc)) +
  geom_point(size = 2, shape = 21)
ath_gg + geom_point(aes(y = 0), colour = "red") +
  geom_segment(aes(xend = weight, yend = 0), linetype = "dashed")
```


```{r}
reg1 = lm(disc ~ weight, data = athletes)
a1 = reg1$coefficients[1] # intercept
b1 = reg1$coefficients[2] # slope
pline1 = ath_gg + geom_abline(intercept = a1, slope = b1,
    col = "blue", linewidth = 1.5)
pline1 + geom_segment(aes(xend = weight, yend = reg1$fitted),
    colour = "red", arrow = arrow(length = unit(0.15, "cm")))
```


```{r}
reg2 = lm(weight ~ disc, data = athletes)
a2 = reg2$coefficients[1] # intercept
b2 = reg2$coefficients[2] # slope
pline2 = ath_gg + geom_abline(intercept = -a2/b2, slope = 1/b2,
    col = "darkgreen", linewidth = 1.5)
pline2 + geom_segment(aes(xend=reg2$fitted, yend=disc),
    colour = "orange", arrow = arrow(length = unit(0.15, "cm")))
```

```{r}
var(athletes$weight) + var(reg1$fitted)
```


```{r}
stopifnot(all.equal(c(var(athletes$weight), var(athletes$disc)), c(1,1)))
xy = cbind(athletes$disc, athletes$weight)
svda = svd(xy)
pc = xy %*% svda$v[, 1] %*% t(svda$v[, 1])
bp = svda$v[2, 1] / svda$v[1, 1]
ap = mean(pc[, 2]) - bp * mean(pc[, 1])
ath_gg + geom_segment(xend = pc[, 1], yend = pc[, 2]) +
  geom_abline(intercept = ap, slope = bp, col = "purple", linewidth = 1.5)
```

```{r}
pline1 + geom_segment(aes(xend = weight, yend = reg1$fitted), colour = "blue", alpha = 0.35) +
  geom_abline(intercept = -a2/b2, slope = 1/b2, col = "darkgreen", linewidth = 1.5, alpha = 0.8) +
  geom_segment(aes(xend = reg2$fitted, yend = disc), colour = "orange", alpha = 0.35) +
  geom_abline(intercept = ap, slope = bp, col = "purple", linewidth = 1.5, alpha = 0.8) +
  geom_segment(xend = pc[, 1], yend = pc[, 2], colour = "purple", alpha = 0.35) + coord_fixed()
```


```{r}
apply(pc, 2, var)
sum(apply(pc, 2, var))
```


    'SVD-mosaicXplot1.png', 'SVD-mosaicXplot2.png', 'SVD-mosaicXplot3.png')))

```{r}
.savedopt = options(digits = 3)
X = matrix(c(780,  75, 540,
             936,  90, 648,
            1300, 125, 900,
             728,  70, 504), nrow = 3)
u = c(0.8196, 0.0788, 0.5674)
v = c(0.4053, 0.4863, 0.6754, 0.3782)
s1 = 2348.2
sum(u^2)
sum(v^2)
s1 * u %*% t(v)
X - s1 * u %*% t(v)
options(.savedopt)
```


```{r}
svd(X)$u[, 1]
svd(X)$v[, 1]
sum(svd(X)$u[, 1]^2)
sum(svd(X)$v[, 1]^2)
svd(X)$d
```

```{r}
Xtwo = matrix(c(12.5, 35.0, 25.0, 25, 9, 14, 26, 18, 16, 21, 49, 32,
       18, 28, 52, 36, 18, 10.5, 64.5, 36), ncol = 4, byrow = TRUE)
USV = svd(Xtwo)
names(USV)
USV$d
```

```{r}
Xtwo - USV$d[1] * USV$u[, 1] %*% t(USV$v[, 1])
Xtwo - USV$d[1] * USV$u[, 1] %*% t(USV$v[, 1]) -
       USV$d[2] * USV$u[, 2] %*% t(USV$v[, 2])

```

```{r}
stopifnot(max(abs(
Xtwo - USV$d[1] * USV$u[, 1] %*% t(USV$v[, 1]) -
       USV$d[2] * USV$u[, 2] %*% t(USV$v[, 2]))) < 1e-12,
max(abs(USV$d[3:4])) < 1e-13)
```

```{r}
t(USV$u) %*% USV$u
t(USV$v) %*% USV$v

```


```{r}
turtles.svd = svd(scaledTurtles)
turtles.svd$d
turtles.svd$v
dim(turtles.svd$u)
```

```{r}
sum(turtles.svd$v[,1]^2)
sum(turtles.svd$d^2) / 47
```

```{r}
stopifnot(max(abs(turtles.svd$v[,1]^2 - 1/3)) < 0.01)
US = turtles.svd$u[,1, drop = FALSE] %*% turtles.svd$d[1]
XV = scaledTurtles %*% turtles.svd$v[, 1, drop = FALSE]
max(abs(US-XV))
stopifnot(max(abs(US-XV)) < 1e-9)

```

```{r}
svda$v[,1]
```


```{r}
ppdf = tibble::tibble(PC1n = -svda$u[, 1] * svda$d[1],
              PC2n =  svda$u[, 2] * svda$d[2])
gg = ggplot(ppdf, aes(x = PC1n, y = PC2n)) + 
    geom_point() + 
    geom_hline(yintercept = 0, color = "purple", linewidth = 1.5, alpha = 0.5) +
    xlab("PC1 ")+ ylab("PC2") + xlim(-3.5, 2.7) + ylim(-2, 2) + coord_fixed()
gg + geom_point(aes(x = PC1n, y = 0), color = "red") +
     geom_segment(aes(xend = PC1n, yend = 0), color = "red") 
gg + geom_point(aes(x = 0, y = PC2n), color = "blue") +
     geom_segment(aes(yend = PC2n, xend = 0), color = "blue") +
     geom_vline(xintercept = 0, color = "skyblue", linewidth = 1.5, alpha = 0.5) 

```

```{r}
sum(ppdf$PC2n^2) 
svda$d[2]^2
stopifnot(abs(sum(ppdf$PC2n^2) - svda$d[2]^2)<1e-9)

```

```{r}
mean(ppdf$PC2n) 
var(ppdf$PC2n) * (nrow(ppdf)-1)
stopifnot(abs(var(ppdf$PC2n) * (nrow(ppdf)-1) - svda$d[2]^2) < 1e-9, abs(mean(ppdf$PC2n)) < 1e-9)

```


```{r}
var(ppdf$PC1n) 
var(ppdf$PC2n) 
stopifnot(var(ppdf$PC1n) > var(ppdf$PC2n))

```


```{r}
sd(ppdf$PC1n) / sd(ppdf$PC2n)
svda$d[1] / svda$d[2]
stopifnot(sd(ppdf$PC1n) / sd(ppdf$PC2n) - svda$d[1] / svda$d[2] < 1e-9)

```


```{r}
cor(scaledTurtles)
pcaturtles = princomp(scaledTurtles)
pcaturtles

```

library("factoextra")
fviz_eig(pcaturtles, geom = "bar", bar_width = 0.4) + ggtitle("")

```{r, eval=FALSE}
svd(scaledTurtles)$v[, 1]
prcomp(turtles[, -1])$rotation[, 1]
princomp(scaledTurtles)$loadings[, 1]
library("ade4")
dudi.pca(turtles[, -1], nf = 2, scannf = FALSE)$c1[, 1]

```

```{r}
res = princomp(scaledTurtles)
PC1 = scaledTurtles %*% res$loadings[,1]
sd1 = sqrt(mean(res$scores[, 1]^2))

```


## -----------------------------------------------------------------------------
fviz_pca_biplot(pcaturtles, label = "var", habillage = turtles[, 1]) +
  ggtitle("")

## -----------------------------------------------------------------------------
pcadudit = dudi.pca(scaledTurtles, nf = 2, scannf = FALSE)
apply(pcadudit$li, 2, function(x) sum(x^2)/48)
pcadudit$eig

## -----------------------------------------------------------------------------
fviz_pca_var(pcaturtles, col.circle = "black") + ggtitle("") +
  xlim(c(-1.2, 1.2)) + ylim(c(-1.2, 1.2))

## -----------------------------------------------------------------------------
svd(scaledTurtles)$d/pcaturtles$sdev
sqrt(47)

## -----------------------------------------------------------------------------
cor(athletes) %>% round(1)

## -----------------------------------------------------------------------------
pca.ath = dudi.pca(athletes, scannf = FALSE)
pca.ath$eig
fviz_eig(pca.ath, geom = "bar", bar_width = 0.3) + ggtitle("")

## -----------------------------------------------------------------------------
fviz_pca_var(pca.ath, col.circle = "black") + ggtitle("")

## -----------------------------------------------------------------------------
athletes[, c(1, 5, 6, 10)] = -athletes[, c(1, 5, 6, 10)]
cor(athletes) %>% round(1)
pcan.ath = dudi.pca(athletes, nf = 2, scannf = FALSE)
pcan.ath$eig

## -----------------------------------------------------------------------------
fviz_pca_var(pcan.ath, col.circle="black") + ggtitle("")

## -----------------------------------------------------------------------------
fviz_pca_ind(pcan.ath) + ggtitle("") + ylim(c(-2.5,5.7))

## -----------------------------------------------------------------------------
data("olympic", package = "ade4")
olympic$score

## -----------------------------------------------------------------------------
p = ggplot(tibble(pc1 = pcan.ath$li[, 1], score = olympic$score, id = rownames(athletes)),
   aes(x = score, y = pc1, label = id)) + geom_text()
p + stat_smooth(method = "lm", se = FALSE)

## -----------------------------------------------------------------------------
load("../data/screep7.RData")
pcaS7 = dudi.pca(screep7, scannf = FALSE)
fviz_eig(pcaS7,geom="bar",bar_width=0.5) + ggtitle("")
#problem with dudi and prcomp eigenvalues
#prcomp does not scale by default, dudi.pca does
#fviz_eig(pcaS7,geom="bar",width=0.3)
#p7=prcomp(screep7,scale= TRUE)
#p7$sdev^2
#plot(p7)

## -----------------------------------------------------------------------------
pcaMsig3 = dudi.pca(Msig3transp, center = TRUE, scale = TRUE,
                    scannf = FALSE, nf = 4)
fviz_screeplot(pcaMsig3) + ggtitle("")

## -----------------------------------------------------------------------------
ids = rownames(Msig3transp)
celltypes = factor(substr(ids, 7, 9))
status = factor(substr(ids, 1, 3))
table(celltypes)
cbind(pcaMsig3$li, tibble(Cluster = celltypes, sample = ids)) %>%
ggplot(aes(x = Axis1, y = Axis2)) +
  geom_point(aes(color = Cluster), size = 5) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_vline(xintercept = 0, linetype = 2) +
  scale_color_discrete(name = "Cluster") + coord_fixed()

## -----------------------------------------------------------------------------
## # For the record:
## library("xcms")
## cdfpath = system.file("cdf", package = "faahKO")
## cdffiles = list.files(cdfpath, recursive = TRUE, full = TRUE)
## xset = xcmsSet(cdffiles)
## xset2 = group(xset)
## xset2 = retcor(xset2)
## xset2 = group(xset2, bw = 10)
## xset3 = fillPeaks(xset2)
## gt = groups(xset3)
## mat1 = groupval(xset3, value = "into")

## -----------------------------------------------------------------------------
load("../data/mat1xcms.RData")
dim(mat1)
pcamat1 = dudi.pca(t(mat1), scannf = FALSE, nf = 3)
fviz_eig(pcamat1, geom = "bar", bar_width = 0.7) + ggtitle("")

## -----------------------------------------------------------------------------
dfmat1 = cbind(pcamat1$li, tibble(
    label = rownames(pcamat1$li),
    number = substr(label, 3, 4),
    type = factor(substr(label, 1, 2))))
pcsplot = ggplot(dfmat1,
  aes(x=Axis1, y=Axis2, label=label, group=number, colour=type)) +
 geom_text(size = 4, vjust = -0.5)+ geom_point(size = 3)+ylim(c(-18,19))
pcsplot + geom_hline(yintercept = 0, linetype = 2) +
  geom_vline(xintercept = 0, linetype = 2)

## -----------------------------------------------------------------------------
pcsplot + geom_line(colour = "red")

## -----------------------------------------------------------------------------
library("pheatmap")
load("../data/wine.RData")
load("../data/wineClass.RData")
wine[1:2, 1:7]
pheatmap(1 - cor(wine), treeheight_row = 0.2)

## -----------------------------------------------------------------------------
winePCAd = dudi.pca(wine, scannf=FALSE)
table(wine.class)
fviz_pca_biplot(winePCAd, geom = "point", habillage = wine.class,
   col.var = "violet", addEllipses = TRUE, ellipse.level = 0.69) +
   ggtitle("") + coord_fixed()

## -----------------------------------------------------------------------------
data("x", package = "Hiiragi2013")
xwt = x[, x$genotype == "WT"]
sel = order(rowVars(Biobase::exprs(xwt)), decreasing = TRUE)[1:100]
xwt = xwt[sel, ]
tab = table(xwt$sampleGroup)
tab

## -----------------------------------------------------------------------------
xwt$weight = 1 / as.numeric(tab[xwt$sampleGroup])
pcaMouse = dudi.pca(as.data.frame(t(Biobase::exprs(xwt))),
  row.w = xwt$weight,
  center = TRUE, scale = TRUE, nf = 2, scannf = FALSE)
fviz_eig(pcaMouse) + ggtitle("")

## -----------------------------------------------------------------------------
## fviz_pca_ind(pcaMouse, geom = "point", col.ind = xwt$sampleGroup) +
##   ggtitle("") + coord_fixed()

## -----------------------------------------------------------------------------
# TODO: duplicate of the above to avoid overlap; 
# can be removed once Quarto resolves this
fviz_pca_ind(pcaMouse, geom = "point", col.ind = xwt$sampleGroup) +
  ggtitle("") + coord_fixed()

## -----------------------------------------------------------------------------
u = seq(2, 30, by = 2)
v = seq(3, 12, by = 3)
X1 = u %*% t(v)

## -----------------------------------------------------------------------------
Materr = matrix(rnorm(60,1),nrow=15,ncol=4)
X = X1+Materr

## -----------------------------------------------------------------------------
outer(u, v)

## -----------------------------------------------------------------------------
ggplot(data=data.frame(X), aes(x=X1, y=X2, col=X3, size=X4)) + geom_point()

## -----------------------------------------------------------------------------
n = 100
p = 4
Y2 = outer(rnorm(n), rnorm(p)) + outer(rnorm(n), rnorm(p))
head(Y2)
ggplot(data=data.frame(Y2), aes(x=X1, y=X2, col=X3, size=X4)) + geom_point()

## -----------------------------------------------------------------------------
svd(Y2)$d # two non-zero eigenvalues
Y = Y2 + matrix(rnorm(n*p, sd=0.01),n,p) # add some noise to Y2
svd(Y)$d # four non-zero eigenvalues (but only 2 big ones)

## -----------------------------------------------------------------------------
library("MASS")
mu1 = 1; mu2 = 2; s1=2.5; s2=0.8; rho=0.9;
sigma = matrix(c(s1^2, s1*s2*rho, s1*s2*rho, s2^2),2)
sim2d = data.frame(mvrnorm(50, mu = c(mu1,mu2), Sigma = sigma))
svd(scale(sim2d))$d
svd(scale(sim2d))$v[,1]

## -----------------------------------------------------------------------------
respc = princomp(sim2d)
dfpc  = data.frame(pc1=respc$scores[,1], 
                   pc2=respc$scores[,2])

ggplot(data.frame(sim2d), aes(x=X1,y=X2)) + geom_point()
ggplot(dfpc, aes(x=pc1, y=pc2)) + geom_point() + coord_fixed(2)

## -----------------------------------------------------------------------------
## require(vcd)
## uvec0=c(1,5.2,0.5,3.6)
## n=length(uvec0)-1
## vvec0=c(1,1.5,1.8,2.5,1.4)
## p=length(vvec0)-1
## rankone=function(uvec=uvec0,vvec=vvec0,factr=100){
## Xout=uvec%*%t(vvec)*factr
## n=length(uvec)
## p=length(vvec)
## dimnames(Xout)=list(U=c(" ",paste("u",1:(n-1),sep="")),V=c(" ",paste("v",1:(p-1),sep="")))
## return(Xout)
## }
## 
## Xmat=rankone()
## 
## makelabels=function(Xout=Xmat,step=0){
## tab=as.table(Xout)
## n=nrow(Xout)
## p=ncol(Xout)
## tab[1,1]=NA
## if (step==0){
##   tab[1,]=rep(NA,p)
## tab[,1]=rep(NA,n)
##   }
##   return(tab)
## }
## 
## tabf0=makelabels(Xmat,step=0)
## tabf1=makelabels(Xmat,step=1)
## Xsub=Xmat[-1,-1]
## dimnames(Xsub)=NULL
## rowns=paste("$x_{",1:n,".}$",sep="")
## colns=paste("$x_{.",1:p,"}$",sep="")
## dimnames(Xsub)=list(rowns,colns)
## require(xtable)
## xtab=xtable(Xsub,floating=FALSE,digits=0)
## xtab

## -----------------------------------------------------------------------------
## np = prod(dim(Xmat))
## rainbow_colors = colorspace::rainbow_hcl(np, c = 50, l = 70, start = 30, end = 360*(np-1)/np)
## mycols = rainbow_colors[1:np]
## mycols[c(1:(n+1), seq(n+2,(n+1)*(p+1),n+1))] = "white"

## -----------------------------------------------------------------------------
## Xmat0=Xmat
## #dimnames(Xmat0)=NULL
## mosaic(Xmat,pop=FALSE,  gp = gpar(fill=matrix(mycols,ncol=p+1),fontsize=14))
## #tabf0
## #tabf1
## labeling_cells(text = tabf0,clip=FALSE,gp_labels = gpar(fontsize = 14))(tabf0)
## #labeling_border(labels =FALSE)

## -----------------------------------------------------------------------------
## tabf2=tabf1
## tabf2[-1,1]=0.1*tabf1[-1,1]
## tabf2[1,-1]=0.1*tabf1[1,-1]
## mosaic(Xmat,pop=FALSE,  gp = gpar(fill=matrix(mycols,ncol=p+1),fontsize=14))
## labeling_cells(text = tabf2,clip=FALSE,gp_labels = gpar(fontsize = 14))(tabf2)

## -----------------------------------------------------------------------------
## tabf3=tabf1
## tabf3[-1,1]=0.05*tabf1[-1,1]
## tabf3[1,-1]=0.2*tabf1[1,-1]
## mosaic(Xmat,pop=FALSE,  gp = gpar(fill=matrix(mycols,ncol=p+1),fontsize=14))
## labeling_cells(text = tabf3,clip=FALSE,gp_labels = gpar(fontsize = 14))(tabf3)

## -----------------------------------------------------------------------------
## USV=svd(Xsub)
## tabusv=tabf0
## tabusv[1,]=c(round(USV$d[1],1),round(USV$v[,1],1))
## tabusv[-1,1]=round(USV$u[,1],1)
## #dimnames(tabusv)[[1]][1]="s1"
## mosaic(Xmat,pop=FALSE,  gp = gpar(fill=matrix(mycols,ncol=5),fontsize = 14))
## labeling_cells(text = tabusv,clip=FALSE,gp_labels = gpar(fontsize = 14))(tabusv)

## -----------------------------------------------------------------------------
## u2=c(5,2,3,4,1.5)
## v2=c(2,7,3,4)
## u3=c(0.5,1,2,2,3)
## v3=c(1,0,4,2)
## Xmat2=rankone(c(1,u2),c(1,v2),1)
## Xmat3=rankone(c(1,u3),c(1,v3),5)
## Xmat4=Xmat2+Xmat3
## tab3f0=makelabels(Xmat4,step=0)
## n=5;p=4;
## mycols[c(1:(n+1),seq(n+2,(n+1)*(p+1),n+1))]="white"
## mosaic(Xmat4,pop=FALSE,  gp = gpar(fill=matrix(mycols,ncol=p+1),fontsize=14))
## labeling_cells(text = tab3f0,clip=FALSE,gp_labels = gpar(fontsize = 14))(tab3f0)

## -----------------------------------------------------------------------------
## tab3f1=makelabels(Xmat3,step=1)
## mosaic(Xmat3,pop=FALSE,  gp = gpar(fill=matrix(mycols,ncol=p+1),fontsize=14))
## labeling_cells(text = tab3f1,clip=FALSE,gp_labels = gpar(fontsize = 14))(tab3f1)

## -----------------------------------------------------------------------------
## tab2f1=makelabels(Xmat2,step=1)
## mosaic(Xmat2,pop=FALSE,  gp = gpar(fill=matrix(mycols,ncol=p+1),fontsize=14))
## labeling_cells(text = tab2f1,clip=FALSE,gp_labels = gpar(fontsize = 14))(tab2f1)

## -----------------------------------------------------------------------------
## Xsub4=Xmat4[-1,-1]
## Xsub3=Xmat3[-1,-1]
## Xsub2=Xmat2[-1,-1]
## svd(Xsub3)
